Towards Generic Parallel Programming in Computer Science Education with Kokkos

### 2020-08-24 src/0.Abstract.tex
Updates for EduHPC2020.

\begin{abstract}\label{chap:abstract}
Parallel patterns, views, and spaces are promising abstractions to capture the programmer's intent as well as the contextual information that can be used by an underlying runtime to efficiently map software to parallel hardware. These abstractions can be valuable in cases where an algorithm must accommodate requirements of code and performance portability across hardware architectures and vendor programming models. Kokkos is a parallel programming model for host- and accelerator architectures that relies on these abstractions and targets these requirements. It consists of a pure C++ interface, a specification, and a programming library. The programming library exposes patterns and types and maps them to an underlying abstract machine model. The abstract machine model offers a generic view of parallel hardware. While Kokkos is gaining popularity in large-scale HPC applications at some DOE laboratories, we believe that the implemented concepts are of interest to a broader audience including academia as they may contribute to a generic, vendor, and architecture-independent education of parallel programming. In this work, we give an insight into the design considerations of this programming model and list important abstractions. Further, we document best practices obtained from giving virtual classes on Kokkos, document attendee feedback, and give pointers to resources that the reader may consider valuable for a lecture on generic parallel programming for students with preexisting knowledge on this matter.
\end{abstract}

>>>>

\begin{abstract}\label{chap:abstract}
Parallel patterns, views, and spaces are promising abstractions to capture the programmer's intent as well as the contextual information that can be used by an underlying runtime to efficiently map software to parallel hardware. These abstractions can be valuable in cases where an algorithm must accommodate requirements of code and performance portability across hardware architectures and vendor programming models. Kokkos is a parallel programming model for host- and accelerator architectures that relies on these abstractions and targets these requirements. It consists of a pure C++ interface, a specification, and a programming library. The programming library exposes patterns and types and maps them to an underlying abstract machine model. The abstract machine model offers a generic view of parallel hardware. While Kokkos is gaining popularity in large-scale HPC applications at some DOE laboratories, we believe that the implemented concepts are of interest to a broader audience including academia as they may contribute to a generic, vendor, and architecture-independent education of parallel programming. In this work, we give an insight into the design considerations of this programming model and list important abstractions. Further, we document best practices obtained from giving virtual classes on Kokkos and give pointers to resources that the reader may consider valuable for a lecture on generic parallel programming for students with preexisting knowledge on this matter.
\end{abstract}

### 2020-08-24 src/1.Introduction.tex
Updates for EduHPC2020.

While message-passing interfaces, OpenMP, OpenCL or CUDA maintain their role in teaching, the plethora of novel parallel programming models~\cite{Chapel}~\cite{UPC++}~\cite{Raja} hints towards an emphasis on productivity in parallel programming and also brings up the valid question whether this strengthens the argument for their inclusion in education.

@@

It is the productivity concern as well as the need for a generic expression of concurrency due to periodic changes of supercomputer architectures and vendors in national projects that inspired the development of Kokkos. The increasing adoption of Kokkos in these projects confirms its usefulness which motivates us to present Kokkos to academia in this work.

@@

The rest of this paper is structured as follows. The next chapter gives an overview of parallel programming concepts and presents the notion of \emph{semantic capture}. Chapter~\ref{chap:kokkosMM} presents the abstract machine model used to derive the required semantic information for performance portability in Kokkos. Chapters~\ref{chap:kokkosPM} and ~\ref{chap:kokkosExample} present the Kokkos programming model and an example application. In Chapter~\ref{chap:backend} we provide insight into a Kokkos backend implementation. Chapters~\ref{chap:related} and ~\ref{chap:teaching} discuss related work, best practices in teaching, and student feedback from tutorials. Lastly Chapter~\ref{chap:conclusion} concludes this work.

>>>>

While message-passing, OpenMP, OpenCL or CUDA maintain their role in teaching, the plethora of novel parallel programming models~\cite{Chapel}~\cite{UPC++}~\cite{Raja} hints towards an emphasis on productivity in parallel programming and also brings up the valid question whether this strengthens the argument for their inclusion in education.

@@

It is the productivity concern as well as the need for a generic expression of concurrency due to periodic changes of supercomputer architectures and their vendors in national projects that inspired the development of Kokkos. The increasing adoption of Kokkos in these projects confirms its usefulness which motivates us to present Kokkos to academia in this work.

@@

The rest of this paper is structured as follows. The next chapter gives an overview of parallel programming concepts and presents the notion of \emph{semantic capture}. Chapter~\ref{chap:kokkosMM} presents the abstract machine model used to derive the required semantic information for performance portability in Kokkos. Chapters~\ref{chap:kokkosPM} and ~\ref{chap:kokkosExample} present the Kokkos programming model and an example application. In Chapter~\ref{chap:backend} we provide insight into a Kokkos backend implementation. Chapters~\ref{chap:related} and ~\ref{chap:teaching} discuss related work and best practices for teaching in online classes. Lastly Chapter~\ref{chap:conclusion} concludes this work.

### 2020-08-24 src/9.Teaching.tex
Updates for EduHPC2020.

The advantage of virtual classes is their possibility of being recording. Recordings can serve as a valuable resource for future reference with quick access links pointing to particular moments in the recording relevant to a topic.

A recent online course\cite{KOKKOS_LECTURE} has been publicly announced through social media and counted approximately 200 attendees. The majority of attendees are familiar with at least one parallel programming model and have a background in scientific computing. The listing in Figure~\ref{fig:lectureOutline} shows the structure of our on-line lecture.

>>>>

The advantage of virtual classes is their possibility of being recording. Recordings can serve as a valuable resource for future reference with quick access links pointing to particular timestamps in the recording relevant to a topic.

A recent online course\cite{KOKKOS_LECTURE} has been publicly announced through social media and counted approximately 200 attendees. The majority of attendees are familiar with at least one parallel programming model and have a background in scientific computing. The listing in Figure~\ref{fig:lectureOutline} shows the structure of our online lecture.
